---
title: "Data and Point Plots"
date: "2020-06-16"
description: "Introduces the distinction between quantitative and categorical variables through their very different appearances in a point plot."
output:
  word_document:
    reference_docx: lesson-style.docx
weight: 100
categories:
  - activities
tags:
  - graphics
  - variables
banner: "img/banners/missing.png"
---

```{r include = FALSE}
SDSdata::sds_setup()
library(LittleApp2)
library(DT)
```

`r SDSdata::word_pdf_links()`

# Orientation

Data can be many things, but one of the most common formats is a data frame, a kind of spreadsheet of rows and columns. We'll work with the data frame (or data set) `Natality_2014` in the Source pakage `Little Apps`, which is based on data  published by the US Centers for Disease Control. `Natality_2014` has 100,000 rows. Each row reports a live birth in the US in 2014. There are dozens of variables, a few of which are shown below.

```{r echo = FALSE}
set.seed(101)
sds_table(sample_n(Natality_2014, size = 10) %>%
               select(sex, dbwt, combgest, me_rot, mager, wic), nrows = 100000, format = "pandoc")
```

It's hard to draw much of a conclusion by looking directly at a large data frame. But a graphical display of data can help.

A *point plot*^[The word "scatterplot" is also used.] is a basic statistical graphic that displays two variables from a data frame. One variable is represented on the vertical axis, another variable on the horizontal axis. Like the following point plot of the baby's weight (in grams) (`dbwt`) and the length (in weeks) of the pregnancy (`combgest`).

```{r echo = FALSE, cache = TRUE, warning = FALSE}
set.seed(101)
gf_point(dbwt ~ combgest, 
         data = sample_n(Natality_2014, size = 5000))

```

# Exercise

Referring to the graph in the previous section ...

1. Find in the graph the dot corresponding to the first row in the data table above, the one for a male baby delivered spontaneously to a 28 year-old mother.
2. Describe the overall pattern shown in the graph as a whole. Use whatever form of description you think is appropriate.
3. Of course, weight differs from one baby to another. In other words, weight *varies*. Describe how much *variation* there is in babies' weight, according to the graph. 
4. Describe how much *variation* there is in gestation length.
5. At which length of gestation are the heaviest babies born?

# Activity

```{r echo = FALSE}
point_plot_app <- list(name = "hello", url = "goodbye")
```


Open the [Regression Little App](`r Regression_app$url`). (See footnote^[<`r Regression_app$rurl`>]).

1. Set the Source package to `Little Apps`, data set to `Natality_2014`.  Choose `dbwt` as the response variable and `combgest` as the explanatory variable. The resulting  plot should look much like the graph seen in the introduction to this lesson.  Change the sample size to $n = 5$ by cliciking on the n=50 icon and choosing n=5. Click on the Graph tab in the top tool bar to see a larger graph.

2. In the "Data" tab in the top tool bar you will see the graph and the data that is in the plot, in data-frame format.

    * For each of the $n=5$ rows of the data frame, find the corresponding point in the graphic.*

3. Change the *explanatory*  variable to `sex`. 

    * For each of the $n=5$ rows of the data frame displayed in  the Data tab, find the corresponding point in the graphic.
    
    * Change $n$ to 500. In the `baby_wt` versus `sex` graph, all the points are lined up in two columns.  
    
    `r question_blank("*Explain why.*", breaks = 3,  length = 3)`


----------------------

Version 0.3, `r Sys.Date()`, Daniel Kaplan, [Word version](`r paste0("/word-versions/",tools::file_path_sans_ext(knitr::current_input()), ".docx")`) 
