---
title: "Types of variables"
date: "2020-06-16"
description: "Introduces the distinction between quantitative and categorical variables, the *range* of a quantitative variable and the *levels* of a categorical variable."
weight: 2
author: "Danny Kaplan"
categories:
  - activities
output:
  word_document:
    reference_docx: lesson-style.docx
  pdf_document: tint::tintPdf
banner: "img/banners/animal-pairs.jpeg"
---


```{r include = FALSE}
SDSdata::sds_setup()
library(LittleApp2)
library(DT)
```

`r SDSdata::word_pdf_links()`

# Orientation

There are two major types of variables:

* **Quantitative**, where the *value* of the variable is a number.
* **Categorical**,  where the value of the variable is one of a set of labels. That is, the value tells which *category* or *group* the row falls into.


Quantitative variables cover an *interval* on the number line. That interval is described by two numbers: the minimum value and the maximum value of the variable.

Quantitative variables also have a *spread*. There are different ways to measure the spread. In this lesson, we'll use the *difference*  between the maximum and minimum value. Note that the *interval* is two numbers while the *spread* is a single number.

The *spread* is said to measure the *variability* in the variable. Other ways to quantify the spread, which we won't use in this lesson, are the *standard deviation* and the *variance*. 

For categorical variables, we don't use the concepts *interval* or *spread*. The possible or allowed values of a categorical variable are called the *levels* of that variable. For example, the levels of a categorical variable describing "commute type" might be: walk, bike, bus, drive, etc. The levels of a categorical variable like "language spoken" might be English, Spanish, Chinese, and so on.

The levels of many categorical variables are *unordered*. This means that the concept of *between* doesn't naturally apply. For instance, there is no  natural way that Spanish is between English and Chinese.

Some categorical variables are *ordinal*, which  is just to say that there is a natural order  to  the levels. An example is a variable recording "opinion," which might have levels Disagree, Neutral, Agree.

# Activity

```{r echo = FALSE}
point_plot_app <- list(name = "hello", url = "goodbye")
```


Open the [Little_App_Spread](`r Spread_app$url`). (See footnote^[<`r Spread_app$url`>]). Make sure you click on Data in the top row. Set the Source package to Little Apps, the Data set to `NHANES2` and set n=200, by clicking on n=50 and pick n=200.

1. Pick a *quantitative response* variable and a *categorical explanatory* variable. You might have to use trial and error to find such variables but once you do, it will be evident in the graph. 

    `r question_blank("*Write down the names of the variables you selected. For the categorical variable, write down each of the levels.*", breaks = 3, length = 0)`

    * Measure the *interval* spanned by the quantitative variable for each group defined by the categorical variable. You can use the measuring stick built into the app. (That is, click on Graph in the top row, then click at a point in the graphic frame and drag the mouse to select a vertical interval. The interval is the information next to your chosen quantitative variable-axis, and the number following the delta symbol is the spread) 
    
    `r question_blank("*Write down the interval spaned by the values of the response variable for each categorical level of the explanatory variable.*", length = 0)`
    
    Level of explanatory variable | Interval spanned by response variable
    -----------------------------------|-----------------------------------
      |  
      |  
      |  
      |  

    * Also, measure and record the *spread* of the quantitative variable for each categorical level. 

2. Using the selector for the explanatory variable, find five more categorical variables in the `NHANES2` data. At least two of your five should be *ordinal* variables.

    `r question_blank("*For each, write down the levels,  and whether the variable is unordered or ordinal.*", breaks = 3)`
    

----------------------

Version 0.3, `r Sys.Date()`, Daniel Kaplan

